---
title: "Machine Learning Assignment"
output: html_document
---

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

### Cross-validation

The "pml_training" data is divided into a training set (60%) and a testing set (40%), while the "pml_testing" is used for validation.

### Expected out-of-sample error

The expected out-of-sample error corresponds to the expected misclassified in the validation data, and it is computed as 1 - accuracy, where the accuracy is obtained when evaluating the testing data on the trained machine learning algorithm.

### Loading libraries and input data

Loading required libraries.

```{r libraries}
library(caret)
library(rpart) 
library(randomForest) 
```

Reading input data and setting to NA all kind of non-valid data.
```{r input_data, cache=TRUE}
pml_training = read.csv( "pml-training.csv",  na.strings = c("NA","#DIV/0!","") )
pml_testing = read.csv( "pml-testing.csv",  na.strings = c("NA","#DIV/0!","") )

set.seed(1234)

```

### Data partitioning

Creating a training and testing sets from "pml_training" data, so that we can train our algorithm on the training set and adjust it using the testing set. Once the best model is selected, results are verified in the validation set from "pml_testing" data.

```{r data_partitionin, cache=TRUE}

inTrain = createDataPartition( y = pml_training$classe, p = 0.6, list = FALSE )

training = pml_training[ inTrain, ];
testing = pml_training[ -inTrain, ];
validation = pml_testing;
```

### Cleaning data

Removing the sample index column in the training data set
```{r cleaning_data_1, cache=TRUE}
training = training[,-1]
```

Removing covariates ith NA entries are filtered out and with very similar variance in the trainging data set.

```{r cleaning_data_2, cache=TRUE}
num_NAs = apply( training , 2, function(x) sum( is.na(x) ) )
valid_cols = num_NAs == 0
training = training[, valid_cols]

nsv = nearZeroVar( training )
training = training[,-nsv]
```

Keeping relevant covariates in the testing and validation data sets. 

```{r cleaning_data_3, cache=TRUE}
testing = testing[, names( training)]
validation = validation[, names(training)[-58] ]
```

### Model testing

Two different machine learning algorithms (decision tree and random forest) are trained on the training data and tested on testing data set.

```{r model_testing_1, cache=TRUE}
modFit1 = rpart( classe ~ ., data = training, method = "class" )
pred1 = predict( modFit1, newdata = testing, type = "class" )
confusionMatrix(pred1, testing$classe )
```

```{r model_testing_2, cache=TRUE}
modFit2 = randomForest(classe ~. , data = training)
pred2 = predict( modFit2, newdata = testing, type = "class" )
confusionMatrix(pred2, testing$classe )
```

Results from the previos seccion indicates that random forest (0.9982) is more accurate than decesion tree (0.8741), so the former is selected.

### Model validation

The selected model is validated on the validation data set. The output is shown below.

```{r model_validation, cache=TRUE}
levels(validation$cvtd_timestamp) = levels(training$cvtd_timestamp)
pred_val = predict( modFit2, newdata = validation, type = "class" )
pred_val
```

